---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Complete end-to-end example

The purpose of this chapter is to inspire you with an example and to expose you to the entire process of creating, publishing, and analyzing your very own website. If you complete the steps described below, you will end up with [this final product](https://r4sites-anatole.netlify.app/). If you like the look and feel of this final product, then all you need to do afterwards is personalize it (e.g. add your own profile image) and author content. These topics are described in the following chapters.

## How it works

The steps to create, publish, and analyze your very own personal website include:

1. Install free software and create free accounts
2. Generate the website using RStudio and the blogdown R package
3. Make some basic edits to placeholder text
4. Integrate Google Analytics to collect user traffic
5. Integrate Formspree to power the contact form
6. Push the website files to GitHub
7. Deploy the live website on Netlify

## Install software

Start off by installing these free, open source tools if you haven’t already. My recommendation is to download the latest versions, which at the time of this writing are R 4.0, RStudio 1.4, Git 2.31.1.

- [R](https://cran.rstudio.com/)
- [RStudio](https://www.rstudio.com/products/rstudio/download/#download)
- [Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git)

## Create accounts {#create-accounts}

Create free accounts with GitHub, Netlify, Google Analytics, and Formspree. Start with GitHub first, then use your GitHub registration to sign up for a Netlify account. To sign up for Google Analytics, click the link below and locate “Start for free” on the landing page.

- [GitHub](https://github.com/)
- [Netlify](https://app.netlify.com/)
- [Google Analytics](https://marketingplatform.google.com/about/analytics/)
- [Formspree](https://formspree.io/register)

## Generate the website {#create-new-blogdown}

[blogdown](https://bookdown.org/yihui/blogdown/) makes it super easy to generate a website that uses [Hugo](https://gohugo.io/). Follow the steps below to create a website on your local machine.

1. Launch RStudio and install the blogdown package by running `install.packages("blogdown")`.

2. Create a new RStudio Project by going to File > New Project > New Directory > Website using blogdown. [Example](https://i.imgur.com/eDg7V2k.png).

3. Using the New Project Wizard, generate a website using the [Hugo Anatole theme](https://themes.gohugo.io/anatole/) by defining `Hugo theme: lxndrblz/anatole` in the configuration. Click Create Project to generate the website. [Example](https://i.imgur.com/lkD1ZlU.png).

4. Confirm that everything is working properly by running `blogdown::serve_site()` in the R console. A local, live, and fully functional preview of your website will appear in the Viewer pane in RStudio. [Example](https://i.imgur.com/oxZgeqM.png).

## Make some edits

You will notice that the default Anatole theme contains placeholder text such as "My Blog" and "Call me Jane" on the homepage below the profile image. Follow the steps below to edit these placeholders.

1. Change "My Blog" on the homepage by navigating to `config/_default/languages.toml` and changing the `title` parameter. [Example](https://github.com/dannymorris/r4sites-demo/blob/master/config/_default/languages.toml#L2).

2. Change "Call me Jane" on the homepage by navigating to `config/_default/params.toml` and editing the `title` parameter. While you're here, feel free to change other parameters including `description`, `profilePicture`, and social media links. [Example](https://github.com/dannymorris/r4sites-demo/blob/master/config/_default/params.toml#L3-L6).

## Integrate Google Analytics

Integrate Google Analytics into your website to collect and analyze user traffic. If you haven't already, be sure to [create a free Google Analytics account](#create-accounts), then follow the steps below.

1. Sign in to Google Analytics and navigate to Admin > Create Account. [Example](https://i.imgur.com/QRTa07k.png)

2. Under Account setup, provide an appropriate name (e.g. my-example-site) and configure the Account Data Sharing Settings to your liking. Click "Next".

3. Under Property setup, provide an appropriate name (e.g. example-site-property). 

4. Under Property setup, click on "Show advanced options" to create a Universal Analytics property. For the Website URL, provide the URL provided by Netlify when you first deployed your website. Click "Next". [Example](https://i.imgur.com/aYeAANl.png)

5. Under About you business, configure the settings to your liking. Click "Next" and accept the terms of service.

6. Locate and copy the Tracking Code by returning the Google Analytics landing page and navigating to the Admin console. Select the Account and Property that were created in the previous steps. Under Property, click on Tracking Info > Tracking Code. Copy the Tracking ID, which should appear in the form of UA-XXXXXXXXX-1. [Example](https://i.imgur.com/qcyaVBD.png)

7. Return to your RStudio Project where you are developing your website. Open the file `config/_default/params.toml` in the main directory and add a `googleAnalytics` parameter with your tracking code, as shown in this [example](https://github.com/dannymorris/r4sites-demo/blob/master/config/_default/params.toml#L16-L17).

## Integrate Formspree

The [contact form](https://r4sites-anatole.netlify.app/contact/) on this demo website is powered by Formspree. Contact forms are a handy feature for collecting user feedback along with user emails. Formspree organizes and stores the responses to your contact form for you on their website. Follow the steps below to create a form "endpoint" and link it to the contact form on the demo website.

- Sign in to [Formspree](https://formspree.io/login).

- Click "New Project" to create a new project for your website. [Example](https://i.imgur.com/jZnrcVL.png).

- Click "New Form" to create a new contact form. [Example](https://i.imgur.com/OYqo3Js.png)

- Copy the form's "endpoint". [Example](https://i.imgur.com/2S2HVO6.png)

- Update the `config/_default/params.toml` file in the main webiste folder to update the `contactFormAction` parameter with your form endpoint, as shown in this [example](https://github.com/dannymorris/r4sites-demo/blob/master/config/_default/params.toml#L18-L19).

## Push the website to GitHub

Once your site is working locally, the next steps are to create a remote GitHub repository for your website, configure a local Git repository, and push your website files from your local environment to the remote repository. Follow these steps to create a remote GitHub repository and sync it to your local Git repository.

1. Create a remote GitHub repository by signing in to [GitHub](https://github.com/) and clicking "New" in the Repositories section. Configure the repository accordingly, then click Create repository. [Example](https://i.imgur.com/LW2AhCr.png)

2. Create a local Git repository and associate it with the GitHub repository created in the previous step. To do this, launch Git from the command line (e.g. Git Bash on Windows, Terminal on Mac/Linux) and running the following Git commands. 


```{bash, eval=F}
# Change the directory to your website's root folder
cd path/to/your/website/files

# initialize local Git repo
git init

# add your website files to the local Git repo
git add --all

# commit your website files to the local Git repo
git commit -m "first commit"

# point the local Git repo to the remote GitHub repo
# change your_username and your_repo to your GitHub username and repo name
git remote add origin https://github.com/your_username/your_repo.git

# push your files to GitHub
git push -u origin master
```

## Deploy the website on Netlify

Now that your website files are stored in your GitHub repository, use Netlify to deploy your website. Netlify will deploy your website directly from your GitHub repository. Any time you push new or edited content to the GitHub repo, Netlify will automatically deploy a new version of your website. This is known as continuous deployment and is key to rapid iteration. Follow these steps to deploy your website on Netlify.

1. Log in to Netlify account using your GitHub registration and click "New site from Git".

2. Select GitHub as the Continuous Deployment provider, then choose the Github repository you configured in the previous steps.

3. Leave the site settings alone and click "Deploy site" to publish your website. [Example](https://i.imgur.com/K4wmIJB.png)

4. After a few seconds, your webiste will be deployed on Netlify. To view your website, return to the Site Overview section and locate the domain name randomly generated by Netlify. [Example](https://i.imgur.com/8TV9NQ1.png). Netlify domains tend to look something like this - https://jovial-roentgen-f2a53c.netlify.app/. To change this, go to Site settings > Change site name.

## Analyze Google Analytics data

[googleAnalyticsR](https://code.markedmondson.me/googleAnalyticsR/) is an R package for querying Google Analytics data. This topic is discussed in more detail in the chapter [Analyzing traffic with Google Analytics](#google-analytics).

The R code below does the following:

- Load R packages for collecting and analyzing Google Analytics data
- Authenticate Google Analytics account
- List accounts and properties within your Google Analytics account, such as the ones associated with your newly created website
- Return a query of users by geographic location 
- Plot a map of users by geographic location

**Note:** The demo website created in the previous sections is not used in the following analysis because it has received no website traffic thus far. My personal website, Abnormal Distributions, is used instead.

```{r, message=F, warning=F, eval=F, echo=F}
# R packages
library(googleAnalyticsR)
library(googleAuthR)
library(dplyr)
library(leaflet)

# Authenticate your Google Analytics account
#options(googleAuthR.scopes.selected = "https://www.googleapis.com/auth/analytics")
#options(googleAuthR.scopes.selected = "https://www.googleapis.com/auth/analytics.edit")
ga_auth(json_file = "googleanalyticsr-auth-key.json")

# List your Google Analytics accounts
ga_accounts <- ga_account_list()

# # Obtain the "view Id" associated with your website account
view_id <- ga_accounts %>%
  filter(accountName == "abndistro", webPropertyName == "Abnormal Distributions") %>%
  pull(viewId)

# Query geographic location of website users
users_location <- google_analytics(
  viewId = "198103217",
  date_range = c("2019-01-01", "2020-12-31"),
  metrics = "users",
  dimensions = c("latitude", "longitude"),
  anti_sample = T
) %>%
  as_tibble() %>%
  mutate_at(vars(latitude, longitude), list(as.numeric))

# Plot map of website users by location
map_plot <- users_location %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~log(users),
    stroke = FALSE,
    fillOpacity = 0.5
 )
```

```{r, message=F, warning=F, eval=F, echo=T}
# R packages
library(googleAnalyticsR)
library(googleAuthR)
library(dplyr)
library(leaflet)

# Authenticate your Google Analytics account
ga_auth()

# List your Google Analytics accounts
ga_accounts <- ga_account_list()

# # Obtain the "view Id" associated with your website account
view_id <- ga_accounts %>%
  filter(accountName == "abndistro", webPropertyName == "Abnormal Distributions") %>%
  pull(viewId)

# Query geographic location of website users
users_location <- google_analytics(
  viewId = "198103217",
  date_range = c("2019-01-01", "2020-12-31"),
  metrics = "users",
  dimensions = c("latitude", "longitude"),
  anti_sample = T
) %>%
  as_tibble() %>%
  mutate_at(vars(latitude, longitude), list(as.numeric))

# Plot map of website users by location
map_plot <- users_location %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~log(users),
    stroke = FALSE,
    fillOpacity = 0.5
 )
```

#### Map of Abnormal Distribtuions users by location {-}

```{r, eval = F}
map_plot
```

## Closing thoughts

The steps to create and publish a website presented in this chapter are discussed in greater detail in the following chapters. Start with the next chapter, [Choosing a Hugo theme](#choose-theme), to learn about choosing a website layout and style that aligns with your vision and goals for your website.

If you like the look and feel of the website created in this chapter, keep using it! You would only need to make a few more edits including the [favicon](#favicon) and any placeholder text. Optional edits include custom images, logos, font styles, color palettes, pages, and language translations. These edits are described in the chapter [Curating the look, feel, and features](#curate-features). If you are interested in jumping straight into writing content (e.g. blog posts), learn how that works in the chapter [Authoring content](#content).
