---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, include = F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, eval=T)
```

# Analyzing traffic with Google Analytics {#google-analytics}

This chapter shows you how to integrate Google Analytics into your website to capture data on your website users including the number of users, the pages they view, the sources from which they arrive at your website, and their geographic locations. This chapter covers the basics of the [googleAnalyticsR](https://code.markedmondson.me/googleAnalyticsR/) R package for querying your Google Analytics data from R. 

If you haven't already, complete the minimum prerequisites listed below:

- [Generate a website](#generate)
- [Deploy your site on Netlify](#deploy)
- [Sign up for Google Analytics](#create-accounts)

## How it works

Once you sign up for Google Analytics, the main objectives are to create an Account and a Property for your website domain name, then obtain a Tracking ID. This tracking code will go in the `config.yaml` file on your website, which in turn will instruct Google Analytics to collect the data on user traffic to your website.

## Setting up Google Analytics {#ga-setup}

Follow these steps to integrate Google Analytics into your website:

1. Sign in to Google Analytics and navigate to Admin > Create Account. [Example](https://i.imgur.com/QRTa07k.png)

2. Under Account setup, provide an appropriate name (e.g. personal-website) and configure the Account Data Sharing Settings to your liking. The defaults should be fine. Click "Next".

3. Under Property setup, provide an appropriate name (e.g. personal-website-property). 

4. Under Property setup, click on "Show advanced options" to create a Universal Analytics property. For the Website URL, provide the URL to your website. Click "Next". [Example](https://i.imgur.com/0jffYnF.png)

5. Under About you business, configure the settings to your liking. Click "Next" and accept the terms of service.

6. Locate the Tracking ID by returning back to the Admin console and selecting the Account and Property that were created in the previous steps. **Important:** Make sure you locate the Universal Analytics property that reads `property name (UA-XXXXXXX-X)` and not the GA4 property. Under Property, click on Tracking Info > Tracking Code ([example](https://i.imgur.com/GtiKfhq.png)). Copy the Tracking ID, which should appear in the form of UA-XXXXXXXXX-1. [Example](https://i.imgur.com/qcyaVBD.png)

7. Return to your RStudio Project where you are developing your website. Open the file `config.yaml` in the main directory and locate the `googleAnalytics:` parameter. Depending on the theme, this parameter may be written or spelled slightly differently. Paste your Google Analytics Tracking ID here, e.g. `googleAnalytics: 'UA-XXXXXXXXX-1'`. [Example](https://i.imgur.com/xgN24ah.png).

8. Push your changes to your GitHub repository to complete the Google Analytics integration.

## Analyzing traffic using googleAnalyticsR

[googleAnalyicsR](https://code.markedmondson.me/googleAnalyticsR/) is an R package for analyzing Google Analytics data stemming from traffic to your website.

### Dimensions and metrics

In order to analyze your Google Analytics data, you need to gain a basic understand of [Dimensions and Metrics](https://support.google.com/analytics/answer/1033861?hl=en).

Dimensions are attributes of your data and Metrics are quantitative measurements. For example, to analyze the count of users on your website in the last 60 days, the Dimension is `date` and the Metric is `users`.

Google Analytics supports hunderds of Dimension-Metric combinations. My recommendation is to browse [Ultimate Google Analytics Dimensions and Metrics List](https://www.ovrdrv.com/blog/ultimate-google-analytics-dimensions-and-metrics-list/) to learn what is available. Also check out the [Practical Tips](https://code.markedmondson.me/googleAnalyticsR/articles/practical-tips.html) page written by the author of the `googleAnalyticsR` package.

### Install googleAnalyticsR

```{r, eval = F}
install.packages("googleAnalyticsR", dependencies = TRUE)
```

### R packages for data analytics

```{r}
library(googleAnalyticsR)  ## calling Google Analytics API
library(dplyr)             ## data manipulation
library(plotly)            ## interactive charts
library(DT)                ## interactive tables
library(leaflet)           ## interactive maps
```

### Authenticate your Google Analytics account

```{r, echo=F, eval=T}
googleAnalyticsR::ga_auth(json_file = "googleanalyticsr-auth-key.json")
```

Running `ga_auth()` in the R console will automatically open a page in your web browser and prompt you to authenticate your Google account.

```{r, echo=T, eval=F}
googleAnalyticsR::ga_auth()
```

### List your Google Analytics accounts and properties

Running `ga_account_list()` will list the Google Analytics account that was created in step 1 of [Setting up Google Analytics](#ga-setup).

```{r}
## get your accounts
ga_accounts <- ga_account_list()

ga_accounts
```

### Obtain viewId associated with your website

To send API calls to Google Analytics, you need to obtain the `viewId` associated with the account.

```{r}
view_id <- ga_accounts %>%
  filter(accountName == "abndistro", webPropertyName == "Abnormal Distributions") %>%
  pull(viewId)

print(view_id)
```

### Daily count of website users

To query the count of website users over a given date range, use the following Dimension and Metric.

- Dimension: `date`
- Metric: `users`

```{r}
daily_users <- google_analytics(
  viewId = view_id,
  date_range = c("2019-07-01", "2019-12-31"),
  metrics = "users",
  dimensions = "date"
) %>%
  as_tibble()

plot_ly(
  data = daily_users,
  x = ~date,
  y = ~users,
  mode = "lines"
) %>%
  layout(title = "Daily count of website between June 2019 and December 2019")
```

### Users by device type

To understand which device types your users are using to view your website, use the following Dimension and Metric.

- Dimension: `deviceCategory`
- Metric: `users`


```{r}
users_by_device_type <- google_analytics(
  viewId = view_id,
  date_range = c("2019-01-01", "2019-12-31"),
  metrics = "users",
  dimensions = "deviceCategory"
) %>%
  as_tibble()

plot_ly(
  data = users_by_device_type,
  x = ~deviceCategory,
  y = ~users,
  type = "bar"
) %>%
  layout(title = "Users by device type")
```

### Page views by page

To measure the number of page views to each of your website pages, use the following Dimension and Metric.

- Dimension: `pageTitle`
- Metric: `pageviews`


```{r}
page_views <- google_analytics(
  viewId = view_id,
  date_range = c("2019-01-01", "2020-12-31"),
  metrics = "pageviews",
  dimensions = "pageTitle"
) %>%
  as_tibble()

page_views %>%
  arrange(desc(pageviews)) %>%
  DT::datatable()
```

### Users by location {#users-by-location}

To understand the geographical areas from which your webiste users view your webiste, use the following Dimension sand Metric.

- Dimensions: `latitude`, `longitude`
- Metric: `users`


```{r}
users_location <- google_analytics(
  viewId = view_id,
  date_range = c("2019-01-01", "2020-12-31"),
  metrics = "users",
  dimensions = c("latitude", "longitude"),
  anti_sample = T
) %>%
  as_tibble() %>%
  mutate_at(vars(latitude, longitude), list(as.numeric))

users_location %>%
  leaflet() %>% 
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~log(users),
    stroke = FALSE, 
    fillOpacity = 0.5
  )
```

### Avoiding data sampling

The `google_analytics()` function has an argument named `anti_sampling`, which is set to `TRUE` by default. This means that queries which return more than 1,000 rows of data will be sampled to avoid putting strain on the Google Analytics API. To avoid this default sampling strategy, set `anti_sampling = FALSE` when making API calls via the `google_analytics()` function.

## Closing thoughts

Google Analytics is a complex topic which cannot be covered in its entirety in this book. For most, having a basic understanding of the available [Google Analytics Dimensions and Metrics](https://www.ovrdrv.com/blog/ultimate-google-analytics-dimensions-and-metrics-list/) should provide a solid foundation for interesting data analytics.


